---
title: Homework 5
author: Ansel George
fontsize: 11pt
output:
  pdf_document:
    highlight: kate
---

```{r, message=F}
library(ggplot2)

set.seed(10)
```


# Problem 1

*Consider the propagated data error in evaluating function $\sin(x)$, i.e., the
error in the function value due to a perturbation $h$ in the argument $x$.*

## a)

*Estimate the absolute error and the relative error in evaluating $\sin(x)$.*

Using a 1st order approximation for $\sin(x)$:

**Absolute error:**

\begin{align}
f(x+h) &= f(x) + hf^\prime(x) \\
\implies f(x+h) - f(x) &= h f^\prime(x) \\
\implies \mathrm{Abs Err}\big(\sin(x)\big) &= h \cos(x)
\end{align}

**Relative error:**

\begin{align}
\mathrm{Rel Err} &= \frac{\mathrm{Abs Err}}{f(x)} \\
\implies \mathrm{Rel Err}\big(\sin(x)\big) &= \frac{h\cos(x)}{\sin(x)} \\
  &= h\cot(x)
\end{align}

If one does not want to compute error based on a 1st order approximation, one
can instead use trigonometric identities to obtain expressions, though they are
more difficult to interpret.

Perturbing $\sin(x)$ by small value $h$ can be expressed as:

\begin{align}
\sin(x+h) &= \sin(x)\cos(h) + \cos(x)\sin(h)
\end{align}

**Absolute error:**

\begin{align}
\sin(x+h) - \sin(x) = \sin(x) (cos(h) - 1) + \cos(x)\sin(h)
\end{align}

**Relative error:**

\begin{align}
\frac{\sin(x+h) - \sin(x)}{\sin(x)} = cos(h) - 1 + \cot(x)\sin(h)
\end{align}


## b)

*Estimate the condition number for this problem. For what values of the
argument $x$ is this problem highly sensitive?*

The relative condition number for estimating $\sin(x)$ is:

\begin{align}
\frac{\left| \Delta y / y \right|}{\left| \Delta x / x \right|} &= \frac{\left| \big( \sin(x+h) - \sin(x) \big)/\sin(x) \right| }{\left| (x + h - x)/x \right|}
\end{align}

As $h \rightarrow 0$:

\begin{align}
\frac{\left| \Delta y / y \right|}{\left| \Delta x / x \right|} &= \frac{\left| \big( \sin(x+h) - \sin(x) \big)/\sin(x) \right| }{\left| (x + h - x)/x \right|} \\
  &= \frac{\left| \sin(x+h) - \sin(x) \right|}{\left| h \right|} \frac{\left| x \right|}{\left| \sin(x) \right|} \\
  &= \left| \cos(x) \right| \frac{\left| x \right| }{\left| \sin(x) \right|} \\
  &= \left| x \cot(x) \right|
\end{align}

The $\cot(x)$ function, which is the $\frac{\cos(x)}{\sin(x)}$, has poles where
$\sin(x) = 0$, specifically when $x = 0, \pi, 2\pi, 3\pi, \dots$.


# Problem 2

*(Cancellation can be a problem for scientific computations on digital
computers. You can read about it in Heath’s Scientific Computing book section
1.3.9.)*

*Write a program to compute the exponential function $e^x$ using the infinite
series: $e^x = 1 + x + x^2 /2! + x^3 /3! + \dots$.*

```{r}
awfulExp <- function(x, threshold=.0005) {
  val <- rep(0, length(x))

  for (i in 1:length(val)) {
    counter <- 0
    error <- 1
    while(error >= threshold) {
      tmp <- x[i]^counter / factorial(counter)
      counter <- counter + 1
      val[i] <- val[i] + tmp
      error <- abs(tmp / val[i])
    }
  }
  return(val)
}
```


## a)
*Summing in the natural order, what stopping criterion should you use?*

The remainder for a truncated Taylor expansion has the lower bound:

\begin{align}
R_n &= \frac{f^{(n+1)}(\xi)}{(n+1)!}(x-h)^{n+1}
\end{align}

for some $\xi$ such that $h \leq \xi \leq x$. For $f(x) = e^x$ evaluated at
$x=0$, this becomes:

\begin{align}
R_n &= \frac{1}{(n+1)!}h^{n+1}
\end{align}

If the desired error tolerance for the computation is $\epsilon_s$ for a given
$x$, then stop the iterations when $\epsilon_s \geq \epsilon_a$, where
$\epsilon_a$ is approximation error measured as:

\begin{align}
\epsilon_a &= \left| \frac{h^{n+1}/(n+1)!}{1 + h + h^2 /2! + h^3 /3! + \dots + h^{n+1}/(n+1)!} \right|
\end{align}

The numerator is the $n+1$th term in the iteration, and the denominator is the
total sum up to that point. The stopping criterion is when $\epsilon_s \geq
\epsilon_a$.

In practical terms, continue adding higher order terms to the approximation
until $\epsilon_a < \epsilon_s$, meaning that the value for the $n$th term in
the iteration is less than the predefined tolerance.


## b)
*Test your program for $x = \pm 1, \pm 5, \pm 10, \pm 15, \pm 20$, and compare
your results with the built-in exponential function $exp(x)$ in the language
you use.*

Estimate for $e^x$ using the Taylor expansion:

```{r}
awfulExp(1)
awfulExp(-1)

awfulExp(5)
awfulExp(-5)

awfulExp(10)
awfulExp(-10)

awfulExp(15)
awfulExp(-15)

awfulExp(20)
awfulExp(-20)
```

Difference between built-in `exp` function:

```{r}
x <- c(1, -1, 5, -5, 10, -10, 15, -15, 20, -20)
qplot(x, abs(exp(x) - awfulExp(x))/ abs(exp(x)), xlab="x", ylab="relative error")
```


## c)
*Can you use the series in this form to get accurate results for $x<0$? (Hint:
$e^{-x} = 1/e^x$.)*

If the supplied value is negative, cancellation problems can arise from simply
summing across terms from the Taylor expansion because it involves summation of
small positive and negative values that cancel.


## d)
*Can you rearrange the series or regroup the terms in any way to get more
accurate results for $x < 0$?*

One can compute the exponential for the negated value (double negative ->
positive) and then invert the final result.

```{r}
betterExp <- function(x, threshold=.0005) {
  val <- rep(0, length(x))

  sgn <- -1*(x<0) + 1*(x>0)
  x <- abs(x)

  for (i in 1:length(val)) {
    counter <- 0
    error <- 1
    while(abs(error) >= threshold) {
      tmp <- x[i]^counter / factorial(counter)
      counter <- counter + 1
      val[i] <- val[i] + tmp
      error <- abs(tmp / val[i])
    }
  }
  return(val^sgn)
}
```

```{r}
x <- c(1, -1, 5, -5, 10, -10, 15, -15, 20, -20)
qplot(x, abs(exp(x) - betterExp(x))/ abs(exp(x)), xlab="x", ylab="relative error")
```


# Problem 3
*Consider the problem of finding the smallest positive root of the nonlinear
equation:*

\begin{align}
\cos(x) + 1/(1 + e^{-2x}) = 0
\end{align}

*Investigate, both theoretically and empirically, the following iterative
schemes for solving this problem using the starting point $x_0 = 3$:*

*(For each scheme, show that it's indeed an equivalent fixed-point problem,
determine analytically whether it's locally convergent and its expected
convergence rate, and then implement the method to confirm your result.)*

Visualize the function:

```{r}
f <- function(x) {
  return(cos(x) + 1/(1+exp(-2*x)))
}
```

```{r}
x <- seq(0, 5, .01)
qplot(x, f(x), geom="line")
```


## a) $x_{k+1} = \arccos \big( -\frac{1}{1 + e^{-2x_k}} \big)$

Rearranging the equation:

\begin{align}
&\cos(x) + 1/(1 + e^{-2x}) = 0 \\
\implies &\cos(x) = -1/(1 + e^{-2x}) \\
\implies &x = \arccos(-1/(1 + e^{-2x}))
\end{align}

Converting the expression to an iteration yields:
\begin{align}
x_{k+1} &= \arccos(-1/(1 + e^{-2x_k})) \\
  &= g(x_k)
\end{align}

This is a fixed point iteration.

### Theoretical

The derivative of $g(x_k)$ is:

\begin{align}
g^\prime(x) &= \frac{\partial}{\partial x} \arccos\Big(\frac{-1}{1 + e^{-2x_k}}\Big) \\
  &= -\bigg( 1 - \frac{1}{(1 + e^{-2x_k})^2} \bigg)^{-\frac{1}{2}}  \bigg(\frac{-2e^{-2x_k}}{(1 + e^{-2x_k})^2}\bigg) \\
  &= \bigg( \frac{(1 + e^{-2x_k})^2 - 1}{(1 + e^{-2x_k})^2} \bigg)^{-\frac{1}{2}}  \bigg(\frac{2e^{-2x_k}}{(1 + e^{-2x_k})^2}\bigg) \\
  &= \bigg( \frac{1}{(1 + e^{-2x_k})^2 - 1} \bigg)^{\frac{1}{2}}  \bigg(\frac{2e^{-2x_k}}{1 + e^{-2x_k}}\bigg) \\
  &= \bigg( \frac{1}{\big( (1 + e^{-2x_k}) + 1 \big) \big(  (1 + e^{-2x_k}) - 1 \big)} \bigg)^{\frac{1}{2}}  \bigg(\frac{2e^{-2x_k}}{1 + e^{-2x_k}}\bigg) \\
  &= \bigg( \frac{1}{2 + e^{-2x_k}} \bigg)^{\frac{1}{2}} \bigg(\frac{2e^{-x_k}}{1 + e^{-2x_k}}\bigg)
\end{align}

The slope of the derivative is less than 1 for all $x$, meaning that the
iterative algorithm will converge on the root. At $x_0 = 3$, the derivative
evaluates to:

```{r}
(1/(2 + exp(-2*3)))^.5 * (2*exp(-3)/(1+exp(-2*3)))
```

The slope can be visualized as follows:

```{r}
g1 <- function(x) {
  return(acos(-1 / (1+exp(-2*x))))
}

g1prime <- function(x) {
  return((1/(2 + exp(-2*x)))^.5 * (2*exp(-x)/(1+exp(-2*x))))
}
```

```{r}
x <- seq(2.5, 3.5, .001)
qplot(x, g1(x), geom="line", ylab="g(x)")
qplot(x, g1prime(x), geom="line", ylab="dg/dx")
```

Assuming linear convergence and that root is close to the starting point $x_0 =
3$, the convergence rate will approximately be $g^\prime(3) \approx 0.07019197$


### Empirical

```{r}
findRootv1 <- function(x0=3, threshold=0.000005, xtrue=NA) {
  error <- 1
  val <- x0
  counter <- 0
  if (!is.na(xtrue)) {
    print("---True error---")
  }
  while(error >= threshold) {
    tmp <- acos(-1/(1+exp(-2*val)))
    # error <- abs(val - tmp)
    error <- abs( (val-tmp) / tmp )
    val <- tmp
    counter <- counter + 1
    if (!is.na(xtrue)) {
      print(paste("Step ", counter, ": ", abs(val - xtrue) / xtrue, sep=""))
    }
  }
  return(list(root=val, steps=counter))
}
```

```{r}
res1 <- findRootv1(3, xtrue=3.076421)
res1$root
res1$steps
```

```{r}
error <- c(0.00167899317266081, 0.000109460173963121, 7.07247533905157e-06,
           4.10335514846105e-07, 2.30827809323527e-08)
x <- seq(1, length(error))
d <- data.frame(error=error, step=x)
ggplot(d, aes(x, log(error)), log="y") +
  geom_point() + geom_smooth(method='lm', formula=y~x)
```

The error ratios is consistent with linear convergence with rate $\approx
0.0609318$, based on a linear regression on the $\log(error)$ values.


## b) $x_{k+1} = 0.5 \log \big( -\frac{1}{1 + \cos^{-1}(x_k)} \big)$

Rearranging the equation:
\begin{align}
\cos(x) + 1/(1 + e^{-2x}) &= 0 \\
\implies 1 + e^{-2x} &= -\cos^{-1}(x) \\
\implies e^{-2x} &= -\cos^{-1}(x) - 1 \\
\implies -2x &= \log\big( -\cos^{-1}(x) - 1 \big) \\
\implies x &= -.5 \log\big( -\cos^{-1}(x) - 1 \big) \\
\implies x &= .5 \log\bigg( \frac{-1}{\cos^{-1}(x) + 1} \bigg)
\end{align}

Converting the expression to an iteration yields:
\begin{align}
  x_{k+1} &= .5 \log\bigg( \frac{-1}{\cos^{-1}(x_k) + 1} \bigg) \\
  &= g(x_k)
\end{align}

### Theoretical

The derivative of $g(x_k)$ is:

\begin{align}
g^\prime(x) &= \frac{d}{d x} 0.5 \log \bigg( -\frac{1}{1 + \cos^{-1}(x_k)} \bigg) \\
  &= \frac{d}{d x} -0.5 \log \big( -1 - \cos^{-1}(x_k) \big) \\
  &= -0.5 \bigg( \frac{1}{\cos^{-1}(x) + 1} \bigg) \tan(x)\sec(x) \\
  &= -0.5 \bigg( \frac{\tan(x)\sec(x)}{\sec(x) + 1} \bigg)
\end{align}

The derivative has a pole when the secant is equal to -1, when $x=\pi$. The
starting point, $x_0=3$, is close to this pole, and evaluating the derivative
that that point yields:

```{r}
-.5*(tan(3)/cos(3) / (1/cos(3) + 1))
```

The magnitude of the slope is much larger than 1, so the iterative algorithm at
that point will diverge.

```{r}
g2 <- function(x) {
  return(.5 * log( -1 / (1 + 1/cos(x))))
}

g2prime <- function(x) {
  return(-.5*(tan(x)/cos(x) / (1/cos(x) + 1)))
}
```

```{r}
x <- seq(2.5, 3.5, .005)
qplot(x, g2(x), ylab="g(x)", geom="line")
qplot(x, g2prime(x), ylab="dg/dx", geom="line")
```

### Empirical

```{r}
findRootv2 <- function(x0=3, threshold=0.000005, xtrue=NA) {
  error <- 1
  val <- x0
  counter <- 0
  if (!is.na(xtrue)) {
    print("---True error---")
  }
  while(error >= threshold) {
    tmp <- .5 * log( -1 / (1 + 1/cos(val)) )
    error <- abs( (val-tmp) / tmp )
    val <- tmp
    counter <- counter + 1
    if (!is.na(xtrue)) {
      print(paste("Step ", counter, ": ", abs(val - xtrue) / xtrue, sep=""))
    }
  }
  return(list(root=val, steps=counter))
}
```

```{r, error=FALSE, ignore=TRUE}
# res2 <- findRootv2(3)
# res2$root
# res2$steps
```


## c) Newton's Method

The derivative of the nonlinear equation is:

\begin{align}
f^\prime(x) &= -\sin(x) - \frac{1}{(1 + e^{-2x})^2}(-2e^{-2x}) \\
  &= -\sin(x) + \frac{2e^{-2x}}{(1 + e^{-2x})^2}
\end{align}

The second derivative is:

\begin{align}
f^{\prime\prime}(x) &= -\cos(x) + \frac{-(1 + e^{-2x})^2 4e^{-2x} + 2e^{-2x}2(1 + e^{-2x})2e^{-2x} }{(1 + e^{-2x})^4} \\
  &= -\cos(x) + \frac{-(1 + e^{-2x}) 4e^{-2x} + 8e^{-2x}e^{-2x} }{(1 + e^{-2x})^3} \\
  &= -\cos(x) + \frac{-4e^{-2x} - 4e^{-4x} + 8e^{-4x}}{(1 + e^{-2x})^3} \\
  &= -\cos(x) + \frac{-4e^{-2x} (1 - e^{-2x})}{(1 + e^{-2x})^3} \\
\end{align}

Newton's method, which is an iterative algorithm of the form $x_{k+1} =
x_k - \frac{f(x_k)}{f^\prime(x_k)}$.


### Theoretical

```{r}
f <- function(x) {
  return(cos(x) + 1/(1+exp(-2*x)))
}

fprime <- function(x) {
  return(-sin(x) + 2*exp(-2*x)/(1+exp(-2*x))^2)
}

fprimeprime <- function(x) {
  return(-cos(x) - 4*exp(-2*x)*(1-exp(-2*x))/(1+exp(-2*x))^3)
}
```

```{r}
x <- seq(2.5, 3.5, .005)
qplot(x, f(x), ylab="f(x)", geom="line")
qplot(x, fprime(x), ylab="df/dx", geom="line")
```


### Empirical

```{r}
findRootNewton <- function(x0=3, threshold=.000005, xtrue=NA) {
  error <- 1
  val <- x0
  counter <- 0
  if (!is.na(xtrue)) {
    print("---True error---")
  }
  while(error >= threshold) {
    tmp <- f(val) / fprime(val)
    val <- val - tmp
    error <- abs(tmp/val)
    counter <- counter + 1
    if (!is.na(xtrue)) {
      print(paste("Step ", counter, ": ", abs(val - xtrue) / xtrue, sep=""))
    }
  }
  return(list(root=val, steps=counter))
}
```

```{r}
res3 <- findRootNewton(3, xtrue=3.076421)
res3$root
res3$steps
```

```{r}
error <- c(0.00685651924068932, 0.00087418595598904, 1.82483540740937e-05,
           4.48766664017057e-08, 5.32413324243922e-08)
x <- seq(1, length(error))
d <- data.frame(error=error, step=x)
ggplot(d, aes(x, log(error)), log="y") +
  geom_point() + geom_smooth(method='lm', formula=y~x)
```

Convergence here is inconsistent but appears less than quadratic.
```{r}
-fprimeprime(3)/fprime(3)/2
error[2:length(error)] / (error[1:(length(error)-1)])^2
error[2:length(error)] / (error[1:(length(error)-1)])
```

The error ratio for each iteration and the previous one is not consistent with
a quadratic convergence but more consistent with (but not exactly) linear.


# Problem 4

*Consider the evolution of DNA sequences in multiple species. Suppose the
divergence time between two sequences is $t$, and the rate of substitution is
$\alpha$. Under the Jukes-Canter model of DNA evolution, the probability of a
nucleotide being substituted is $s_t = 3(\frac{1}{4} - \frac{1}{4}e^{-4\alpha
t/3})$, and the probability of a nucleotide being conserved is: $r_t = 1-s_t =
\frac{1}{4} + \frac{3}{4}e^{-4\alpha t/3}$.*

## a)
*Given a pair of DNA sequences $X_1$ and $X_2$, let $t$ be the divergence time
from one sequence to the other. Since $\alpha$ and $t$ are coupled, we can set
$\alpha$ to $1$. Write the likelihood function of $t$ in terms of the number of
nucleotide substitution pairs $n_S$ and the number of conserved nucleotide
pairs $n_C$ assuming a binomial model.*

\begin{align}
L(t; n_S, n_C) &= { n_S + n_C \choose n_S } s_t^{n_S} r_t^{n_C} \\
  &= { n_S + n_C \choose n_S } \bigg(\frac{1}{4}\bigg)^{n_S + n_C} \big(3 - 3 e^{-4t/3} \big)^{n_S} \big(1 + 3 e^{-4t/3} \big)^{n_C} \\
  &\propto \big(3 - 3 e^{-4t/3} \big)^{n_S}  \big(1 + 3 e^{-4t/3} \big)^{n_C}
\end{align}


## b)
*Implement the Newton's method to give the maximum likelihood estimate of $t$
when $n_S = 5$ and $n_C = 25$.*

The maximum likelihood estimate is the value of $t$ that maximizes the value of
the likelihood. This corresponds to points where the derivative of the
likelihood (or log-likelihood) is 0. These roots can be found by Newton's
method.

Let $f(t)$ be proportional to the derivative of $L(t;n_S,n_C)$:

\begin{align}
f(t) &= \frac{d}{dt}\bigg\{ \big(3 - 3 e^{-4t/3} \big)^{n_S}  \big(1 + 3 e^{-4t/3} \big)^{n_C}\bigg\}\\
  &= n_S \big(3 - 3 e^{-4t/3} \big)^{n_S-1} 4e^{-4t/3} \big(1+3e^{-4t/3} \big)^{n_C} + \big(3-3e^{-4t/3} \big)^{n_S} n_C\big(1+3e^{-4t/3} \big)^{n_C-1} -4e^{-4t/3} \\
  &=  4e^{-4t/3} \big(3 - 3 e^{-4t/3} \big)^{n_S-1} \big(1+3e^{-4t/3} \big)^{n_C-1} \Big(-n_C \big(3-3e^{-4t/3} \big) + n_S \big(1+3e^{-4t/3} \big) \Big) \\
  &=  4e^{-4t/3} L\big(t; n_S-1, n_C-1\big) \Big( (n_S-3n_C) + 3(n_C+n_S)e^{-4t/3} \Big) \\
  &=  L\big(t; n_S-1, n_C-1\big) \Big( 4(n_S-3n_C)e^{-4t/3} + 12(n_C+n_S)e^{-8t/3} \Big)
\end{align}

The derivative of this expression $f^\prime(t)$ is:
\begin{align}
f^\prime(t) =& \frac{d}{dt} \bigg\{ L\big(t; n_S-1, n_C-1\big) \Big( 4(n_S-3n_C)e^{-4t/3} + 12(n_C+n_S)e^{-8t/3} \Big) \bigg\} \\
  =& \frac{d}{dt} \bigg\{ L\big(t; n_S-1, n_C-1\big) \bigg\} \Big( 4(n_S-3n_C)e^{-4t/3} + 12(n_C+n_S)e^{-8t/3} \Big) + \\
  &L\big(t; n_S-1, n_C-1\big) \frac{d}{dt} \bigg\{\Big( 4(n_S-3n_C)e^{-4t/3} + 12(n_C+n_S)e^{-8t/3} \Big) \bigg\} \\
  =& f(t; n_S-1, n_C-1) \Big( 4(n_S-3n_C)e^{-4t/3} + 12(n_C+n_S)e^{-8t/3} \Big) + \\
  &L\big(t; n_S-1, n_C-1\big) \Big( \frac{-16}{3}(n_S-3n_C)e^{-4t/3} + \frac{-96}{3}(n_C+n_S)e^{-8t/3} \Big)
\end{align}

Newton's method has the form $x_{k+1} = x_k - \frac{f(x_k)}{f^\prime(x_k)}$.

```{r}
# likelihood of binomial distribution
L <- function(t, ns, nc) {
  val <- (3-3*exp(-4/3*t))^ns * (1+3*exp(-4/3*t))^nc
  return(val)
}

# 1st derivative f(t) of the likelihood
f <- function(t, ns, nc) {
  val <- L(t, ns-1, nc-1) * (4*(ns-3*nc)*exp(-4/3*t) + 12*(nc+ns)*(exp(-8/3*t)))
  return(val)
}

# 2nd derivative of likelihood f'(t)
fprime <- function(t, ns, nc) {
  val <- L(t, ns-1, nc-1) * (-16/3*(ns-3*nc)*exp(-4/3*t) + -96/3*(nc+ns)*(exp(-8/3*t)))
  val <- val + f(t, ns-1, nc-1) * (4*(ns-3*nc)*exp(-4/3*t) + 12*(nc+ns)*(exp(-8/3*t)))
  return(val)
}

# Modified Newton-Raphson root finder that will pass the ns, nc parameters to
# the likelihood functions.
findRootNewton <- function(x0=10, ns=5, nc=25, threshold=.005) {
  error <- 1
  val <- x0
  counter <- 0
  while(error >= threshold) {
    tmp <- f(val, ns, nc) / fprime(val, ns, nc)
    val <- val - tmp
    error <- abs(tmp/val)
    counter <- counter + 1
  }
  return(list(root=val, steps=counter))
}
```

Because the Newton-Raphson method has limitations in that it does not guarantee
which root is found and does can get caught searching for extrema since the
likelihood function tails asymptotically to 0, visually look at the likelihood
to determine where a root of f(t) might be that could correspond to the MLE
estimate.


```{r}
ns <- 5
nc <- 25
t <- seq(0,1,.005)
qplot(t, L(t, ns, nc), geom="line")
qplot(t, f(t, ns, nc), geom="line")
qplot(t, fprime(t, ns, nc), geom="line")
```

There is a single root in $f(t)$ near $0.2$. It corresponds to a negative value
in $f^\prime(t)$, meaning that it is a maximum and therefore a MLE for $L(t)$.

```{r}
x0 <- .2
res <- findRootNewton(x0, ns, nc, threshold=0.000005)
print(paste(res$root))
res$steps
```


# Problem 5

*Implement the Gram-Schmidt process for QR decomposition, including a function
that can check whether the resulting basis vectors are orthonormal. Please give
an example with input and output.*

```{r}
gramschmidtQR <- function(A) {
  m <- dim(A)[1]
  n <- dim(A)[2]
  Q <- matrix(0, nrow=m, ncol=n)
  R <- matrix(0, nrow=n, ncol=n)
  for (j in 1:n) {
    v <- A[,j]
    for (i in 1:j) {
      R[i,j] <- Q[,i] %*% A[,j]
      v <- v - R[i,j] * Q[,i]
    }
    R[j,j] <- norm(v, type="2")
    Q[,j] <- v / R[j,j]
  }
  return(list(Q=Q, R=R))
}
```

```{r}
m <- 10
n <- 8

A <- matrix(rnorm(m*n), nrow=m, ncol=n)
res <- gramschmidtQR(A)

res$Q
res$R

# Check that the Q and R are correct.
all.equal(A, res$Q %*% res$R)
```

To check whether the matrix $Q$ is orthonormal, check that that $Q^T Q = I$,
where $I$ is the identity matrix. The `all.equal` function is used to add
tolerance for rounding error that could make $Q^T Q$ not exactly equal to $I$.

```{r}
isOrthonormal <- function(Q) {
  if (all.equal(diag(dim(A)[2]), t(res$Q) %*% res$Q)) {
    return(TRUE)
  } else {
    return(FALSE)
  }
}
```

```{r}
isOrthonormal(res$Q)
```
