---
title: Homework 6
author: Ansel George
output: 
    pdf_document:
        highlight: kate
fontsize: 11pt
---

```{r, message=F}
library(Rglpk)
library(quadprog)

set.seed(10)
```

\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

# Dantzig selector

\begin{align*}
\min_{\boldsymbol\beta} \norm{\boldsymbol\beta}_1 \textrm{ st. } |\boldsymbol{y}_i - \boldsymbol{x}_i^T \boldsymbol\beta | \leq \lambda \textrm{ for } i=1,\dots,n
\end{align*}

## Recast the problem into a linear programming.

Define $\boldsymbol{e} > |\boldsymbol{\beta}|$, where $\boldsymbol{e} \in
\mathbb{R}^p$ and $\boldsymbol{e} \geq \boldsymbol{0}$.

The inequalities and constraints can be formulated as the following system of
inequalities.

\begin{align*}
\boldsymbol{\beta} &\leq \boldsymbol{e} \\
-\boldsymbol{\beta} &\leq \boldsymbol{e} \\
\boldsymbol{y} - X\boldsymbol\beta &\leq \lambda \boldsymbol1 \\
-\boldsymbol{y} + X\boldsymbol\beta &\leq \lambda \boldsymbol1
\end{align*}

Rearranging:

\begin{align*}
\boldsymbol{e} - \boldsymbol{\beta} &\geq 0\\
\boldsymbol{e} + \boldsymbol{\beta} &\geq 0\\
X\boldsymbol\beta &\geq  \boldsymbol{y} - \lambda \boldsymbol1\\
- X\boldsymbol\beta &\geq - \lambda \boldsymbol1 - \boldsymbol{y}  \\
\end{align*}

This can be written in matrix form as:

\[
\begin{bmatrix}
I & -I \\
I & I \\
0 & X \\
0 & -X \\
\end{bmatrix}
\begin{bmatrix}
\boldsymbol{e} \\
\boldsymbol{\beta} \\
\end{bmatrix}
=
\begin{bmatrix}
0 \\
0 \\
\boldsymbol{y} - \lambda\boldsymbol{1} \\
-\boldsymbol{y} - \lambda\boldsymbol{1}
\end{bmatrix}
\]

\[
G
\begin{bmatrix}
\boldsymbol{e} \\
\boldsymbol{\beta} \\
\end{bmatrix}
=
\boldsymbol{\gamma}
\]

These equations can then be used to formulate a linear programming problem:

\begin{align*}
\min_{\boldsymbol{e}} \boldsymbol1^T\boldsymbol{e} \textrm{ where }& \boldsymbol{e} > \boldsymbol{0}\textrm{, and} \\
&\boldsymbol{G}
\begin{bmatrix}
\boldsymbol{e} \\
\boldsymbol{\beta} \\
\end{bmatrix}
=
\boldsymbol{\gamma}
\end{align*}

## Use R interface of a linear programming solver (such as `rmosek`, `cplex`,
  `Rglpk`, `clpAPI`).

*Follow the examples in Convex demo.pdf.*


```{r}
n <- 15
p <- 10

lambda <- rep(5, n)
y <- ceiling(runif(n, 0, 2*n))-n
X <- matrix(ceiling(runif(n*p, 0, 2*n))-n, ncol=p, nrow=n)
```


```{r}
obj <- c(rep(1, p), rep(0, p))
G <- matrix(0, nrow=(2*p+2*n), ncol=(2*p))
G[1:p,1:p] <- diag(p)
G[(p+1):(2*p),1:p] <- diag(p)
G[(2*p+1):(2*n+2*p),1:p] <- 0

G[1:p,(p+1):(2*p)] <- diag(p)
G[(p+1):(2*p),(p+1):(2*p)] <- -diag(p)
G[(2*p+1):(n+2*p),(p+1):(2*p)] <- X
G[(n+2*p+1):(2*n+2*p),(p+1):(2*p)] <- -X

direction <- rep(">=", (2*n+2*p))
rhs <- c(rep(0, 2*p), (y-lambda), -(y+lambda))
bounds <- list(lower=list(ind=(p+1):(2*p), val=rep(-Inf, p)))

soln <- Rglpk_solve_LP(obj, G, direction, rhs, bounds=bounds)
e <- soln$solution[1:p]
b <- soln$solution[(p+1):(2*p)]

soln$status      # return status of solver
sum(abs(b))      # 1-norm of b
abs(y - X %*% b) # check that |y - Xb| is less than lambda
```


Again with a different matrix $X$:

```{r}
X <- matrix(ceiling(rnorm(n*p, 0, 5)), ncol=p, nrow=n)
```

```{r}
obj <- c(rep(1, p), rep(0, p))
G <- matrix(0, nrow=(2*p+2*n), ncol=(2*p))
G[1:p,1:p] <- diag(p)
G[(p+1):(2*p),1:p] <- diag(p)
G[(2*p+1):(2*n+2*p),1:p] <- 0

G[1:p,(p+1):(2*p)] <- diag(p)
G[(p+1):(2*p),(p+1):(2*p)] <- -diag(p)
G[(2*p+1):(n+2*p),(p+1):(2*p)] <- X
G[(n+2*p+1):(2*n+2*p),(p+1):(2*p)] <- -X

direction <- rep(">=", (2*n+2*p))
rhs <- c(rep(0, 2*p), (y-lambda), -(y+lambda))
bounds <- list(lower=list(ind=(p+1):(2*p), val=rep(-Inf, p)))

soln <- Rglpk_solve_LP(obj, G, direction, rhs, bounds=bounds)
e <- soln$solution[1:p]
b <- soln$solution[(p+1):(2*p)]

soln$status      # return status of solver
sum(abs(b))      # 1-norm of b
abs(y - X %*% b) # check that |y - Xb| is less than lambda
```


# Markowitz portfolio optimization

*Suppose there are $p$ assets held over a fixed period and let $\beta_j$ denote
the amount invested in the $j$-th asset. Let $R$ be a $p$-variate random
variable denoting the relative price change of each asset. We assume $R$ has
mean $\mu_R$ and covariance $\Sigma_R$.*

\begin{align*}
\min_{\boldsymbol\beta}\boldsymbol{\beta}^T \Sigma_R \boldsymbol\beta \textrm{ st. }& \boldsymbol\beta^T \boldsymbol\mu_R \geq 1 \\
  &\boldsymbol1^T \boldsymbol\beta = 1 \\
  &\boldsymbol\beta \geq \boldsymbol0
\end{align*}

*Can you implement Markowitz portfolio optimization using R interface of a
quadratic programming solver?*

The goal is to construct a portfolio with a maximum in return and minimum in
variance in return.

```{r}
n <- 8
p <- 8

X <- matrix(rnorm(n*n), ncol=n, nrow=n)
X <- X %*% t(X)
mu <- runif(n, 0, 2)
```

```{r}
d <- rep(0, p)
A <- rbind(rep(1, p), mu)
A <- rbind(A, diag(p))
rhs <- c(1, min(1, max(mu)), rep(0, p))
soln <- solve.QP (2*X, d, t(A), rhs, meq=1)
b <- soln$solution

sum(b)        # should sum to 1
b %*% X %*% b # variance in return
b %*% mu      # average return
```
